---
title: "Data_Wrangling"
output:
  html_document: default
  pdf_document: default
date: "2026-01-06"
---


```{r}

library(rstatix)
library(tidyverse)
library(pastecs)
library(knitr)
library(epitools)
library(Epi)
library(gtsummary)
library(janitor)
library(scales)     
library(viridis)    
options(scipen=999) 
```


## Question 1: Data Cleaning


```{r}
## Load the first datasets- assignment1_can_path_demographics

data_1 <- read_csv("assignment1_can_path_demographics.csv")
```


```{r}
## Examining data structure- Columns, and data types

glimpse(data_1)
```


```{r}
## Examining data structure- Unique values 

table(data_1$SDC_GENDER)
table(data_1$SDC_INCOME)

```



```{r}
## Identify duplicates- rows

duplicate_rows <- data_1[duplicated(data_1), ]

duplicate_rows

```


```{r}
## Identify duplicates- any

any(duplicated(data_1))
```

### No duplicates, outliers and inconsistent formatting found in data_1


```{r}
## Recoding gender

data_1 <- data_1 %>%
	mutate(gender_recode = case_when(
		SDC_GENDER == 1 ~ "1_Male",
		SDC_GENDER == 2 ~ "2_Female"
    	))
data_1$gender_recode <- as.factor(data_1$gender_recode)
table(data_1$SDC_GENDER, data_1$gender_recode)
```


```{r}
## Recoding income

data_1 <- data_1 %>%
	mutate(income_recode = case_when(
		SDC_INCOME == 1 ~ "1_Less than 10 000 $",
    SDC_INCOME == 2 ~ "2_10 000 $ - 24 999 $",
		SDC_INCOME == 3 ~ "3_25 000 $ - 49 999 $",
		SDC_INCOME == 4 ~ "4_50 000 $ - 74 999 $",
    SDC_INCOME == 5 ~ "5_75 000 $ - 99 999 $",
		SDC_INCOME == 6 ~ "6_100 000 $ - 149 999 $",
    SDC_INCOME == 7 ~ "7_150 000 $ - 199 999 $",
    SDC_INCOME == 8 ~	"8_200 000 $ or more"
	))

data_1$income_recode <- as.factor(data_1$income_recode)
table(data_1$SDC_INCOME, data_1$income_recode)


```

```{r}
## Checking the cleaned data

glimpse(data_1)
```


```{r}
## Load the second dataset- assignment1_can_path_health

data_2 <- read_csv("assignment1_can_path_health.csv")
```


```{r}
## Examining data structure- Columns, and data types

glimpse(data_2)
```

```{r}
## Examining data structure- Unique values 

table(data_2$HS_GEN_HEALTH)
table(data_2$DIS_ASTHMA_EVER)
table(data_2$NUT_VEG_QTY)
table(data_2$NUT_FRUITS_QTY)

## table(data_2$PA_TOTAL_SHORT) ## Too large to display
```



```{r}
## Identify duplicates- rows

duplicate_rows <- data_2[duplicated(data_2), ]
duplicate_rows
```

```{r}
## Identify duplicates- any

any(duplicated(data_2))
```
### No duplicates found in data_2


### Identifing and correcting outliers- NUT_VEG_QTY

#### I am thinking that other approaches such as transformation are not under the scope of this assignment and I have followed the data cleasing approaches included in the course materials.

```{r}

table(data_2$NUT_VEG_QTY)
summary(data_2$NUT_VEG_QTY)
```


```{r}
## Boxplot to see outliers

boxplot_veg <- ggplot(data_2, aes(NUT_VEG_QTY)) + 
              geom_boxplot() + 
              coord_flip()  

plot(boxplot_veg)
```


```{r}
 # Recoding to correct for outliers 

data_2 <- data_2 %>%
          mutate(nut_veg_qty = case_when(
            NUT_VEG_QTY < 1 ~ 1,
            NUT_VEG_QTY > 5 ~ 5,
            TRUE ~ NUT_VEG_QTY
          ))

table (data_2$nut_veg_qty)

summary(data_2$nut_veg_qty)

```

```{r}
## Boxplot to see the outliers after recoding/ correction

boxplot_veg <- ggplot (data_2, aes(nut_veg_qty)) +
               geom_boxplot() +
               coord_flip ()

plot(boxplot_veg)


```

### Identifing and correcting outliers- NUT_FRUITS_QTY

```{r}
summary(data_2$NUT_FRUITS_QTY)
table(data_2$NUT_FRUITS_QTY)

```

```{r}
## Boxplot to see outliers

boxplot_fruit <- ggplot (data_2, aes(NUT_FRUITS_QTY)) +
               geom_boxplot() +
               coord_flip ()

plot(boxplot_fruit)
```


```{r}
 # Recoding to correct for outliers

data_2 <- data_2 %>%
          mutate(nut_fruit_qty = case_when(
            NUT_FRUITS_QTY > 5 ~ 5,
            TRUE ~ NUT_FRUITS_QTY
          ))

table (data_2$nut_fruit_qty)

summary(data_2$nut_fruit_qty)

```

```{r}
#Identify and correct issues- outliers_NUT_VEG_QTY

boxplot_fruit <- ggplot (data_2, aes(nut_fruit_qty)) +
               geom_boxplot() +
               coord_flip ()

plot(boxplot_fruit)
```

### Identify and correct outliers- PA_TOTAL_SHORT

```{r}
summary(data_2$PA_TOTAL_SHORT)
```

```{r}
## Boxplot to see outliers


boxplot_pa <- ggplot (data_2, aes(PA_TOTAL_SHORT)) +
               geom_boxplot() +
               coord_flip ()

plot(boxplot_pa)
```

```{r}
 # Recoding to correct for outliers

data_2 <- data_2 %>%
          mutate(pa_total_short_rec = case_when(
           PA_TOTAL_SHORT > 6000 ~ 6000,
            TRUE ~ PA_TOTAL_SHORT
          ))
```


```{r}
summary(data_2$pa_total_short_rec)
```




```{r}
## Boxplot to see the outliers after recoding/ correction

boxplot_pa_rec <- ggplot (data_2, aes(pa_total_short_rec)) +
               geom_boxplot() +
               coord_flip ()
plot(boxplot_pa_rec)
```

### No inconsistent formatting found in data_2


```{r}
## Recoding general health

data_2 <- data_2 %>%
	mutate(health_recode = case_when(
		HS_GEN_HEALTH == 1 ~ "1_Poor",
		HS_GEN_HEALTH == 2 ~ "2_Fair",
    HS_GEN_HEALTH == 3 ~ "3_Good",
    HS_GEN_HEALTH == 4 ~ "4_Very good",
    HS_GEN_HEALTH == 5 ~ "5_Excellent"
    	))

data_2$health_recode <- as.factor(data_2$health_recode)
table(data_2$HS_GEN_HEALTH, data_2$health_recode)

```


```{r}
## Recoding asthma status

data_2 <- data_2 %>%
	mutate(asthma_recode = case_when(
		DIS_ASTHMA_EVER == 0 ~ "0_Never had asthma",
		DIS_ASTHMA_EVER == 1 ~ "1_Ever had asthma",
    DIS_ASTHMA_EVER == 2 ~ "2_Presumed - Never had asthma"
    	))
data_2$asthma_recode <- as.factor(data_2$asthma_recode)
table(data_2$DIS_ASTHMA_EVER, data_2$asthma_recode)


```

## Question 2: Joining Datasets

```{r}
## Combine the two datasets using left join function

data_all <- left_join(data_1, data_2, by = join_by(ID))

```

```{r}
## Maintaining consistency and completeness- Comparing the datasets

glimpse(data_1)
glimpse(data_2)
glimpse(data_all)

```


```{r}
## Maintaining consistency and completeness- Summary statistic

summary(data_1)
summary(data_2)
summary(data_all)
```

```{r}
## Maintaining consistency and completeness- Checking number of rows only
nrow(data_1)
nrow(data_2)
nrow(data_all)

```


```{r}
## Maintaining consistency and completeness-  Check for duplicates 

duplicate_rows <- data_all[duplicated(data_all), ]
duplicate_rows

any(duplicated(data_all))
```

### Based on the above data quality checks mechanisms, the merged dataset is complete and consistent.  


## Question 3: Descriptive Statistics


```{r}
# Descriptive statistics for  categorical variables 
tbl1 <- data_all %>%
  select(gender_recode, income_recode, health_recode, asthma_recode) %>%
  tbl_summary(
    type = all_categorical() ~ "categorical",
    statistic = all_categorical() ~ "{n} ({p}%)",
    missing = "ifany", 
    missing_text = "Missing",     
    missing_stat = "{n} ({p}%)",
    label = list(
      gender_recode ~ "Gender",
      income_recode ~ "Income Level",
      health_recode ~ "General Health",
      asthma_recode ~ "Asthma Status"
    )
  ) %>%
  add_n() %>%
  bold_labels()

tbl1

```

```{r}
## Calculating key statistics for Physical Activity


summary(data_all$pa_total_short_rec)
```


```{r}
## Calculating key statistics for Physical Activity- with more details


mean_pa <- mean(data_all$pa_total_short_rec, na.rm = TRUE)
median_pa <- median(data_all$pa_total_short_rec, na.rm = TRUE)
sd_pa <- sd(data_all$pa_total_short_rec, na.rm = TRUE)
var_pa <- var(data_all$pa_total_short_rec, na.rm = TRUE)

pa_stats <- data.frame(
  Statistic = c("Mean", "Median", "Standard Deviation", "Variance"),
  Value = c(mean_pa, median_pa, sd_pa, var_pa)
)
kable(pa_stats, caption = "Statistics for PA_TOTAL_SHORT")


```

### Issues identified in physical activity variable: The Mean value= 2309 while the meadian value is 1782 indicating the fact that the distribution is right-skewed. The standard deviation is also high (1982), which could suggest the the high levels of variability/dispersion in physical activity levels among respondents.The Interquartile range is also broad (3732 − 600= 3131) is broad that indicates the wide spread of the data. There is also large missing data-  6763. 


```{r}
## Calculating key statistics for vegetable consumption

summary(data_all$nut_veg_qty)
```

```{r}
## Calculating key statistics for for vegetable consumption- - with more details

mean_veg <- mean(data_all$nut_veg_qty, na.rm = TRUE)
median_veg <- median(data_all$nut_veg_qty, na.rm = TRUE)
sd_veg <- sd(data_all$nut_veg_qty, na.rm = TRUE)
var_veg <- var(data_all$nut_veg_qty, na.rm = TRUE)

veg_stats <- data.frame(
  Statistic = c("Mean", "Median", "Standard Deviation", "Variance"),
  Value = c(mean_veg, median_veg, sd_veg, var_veg)
)
kable(veg_stats, caption = "Statistics for NUT_VEG_QTY")

```

### Issues identified in vegetable consumption variable: The Mean value= 2.67 while the meadian value is 3 indicating the fact that the distribution is slightly left-skewed. There is also large missing data-  2549.


```{r}
## Calculating key statistics for fruit consumption

summary(data_all$nut_fruit_qty)
```


```{r}
## Calculating key statistics for for fruit consumption- - with more details

mean_fruit <- mean(data_all$nut_fruit_qty, na.rm = TRUE)
median_fruit <- median(data_all$nut_fruit_qty, na.rm = TRUE)
sd_fruit <- sd(data_all$nut_fruit_qty, na.rm = TRUE)
var_fruit <- var(data_all$nut_fruit_qty, na.rm = TRUE)

fruit_stats <- data.frame(
  Statistic = c("Mean", "Median", "Standard Deviation", "Variance"),
  Value = c(mean_fruit, median_fruit, sd_fruit, var_fruit)
)
kable(fruit_stats, caption = "Statistics for NUT_FRUITS_QTY")
```


### Issues identified in fruit consumption variable: The Mean value= 2.09 while the meadian value is 2.0 indicating the fact that the data is normally ditributed. There is also large missing data-  2426.



## Question 4: Data Visualization


```{r}
## Bar chart showing income distribution by gender

bar_income <- ggplot(
  data_all |> mutate(gender_recode = fct_explicit_na(gender_recode, na_level = "Missing")),
  aes(income_recode, fill = gender_recode)
) +
  geom_bar(width = 0.75, color = "white", linewidth = 0.3, position = "stack") +
  labs(
    title = "Income Distribution by Gender",
    x     = "Income Categories",
    y     = "Number of respondents",
    fill  = "Gender"
  ) +
  scale_fill_viridis_d(option = "D", end = 0.9) +  # color-blind friendly
  theme_minimal(base_size = 12) +
  theme(
    plot.title        = element_text(face = "bold", size = 13),
    plot.subtitle     = element_text(color = "grey30"),
    axis.text.x       = element_text(angle = 45, hjust = 1),
    panel.grid.major.x= element_blank()
  )

plot(bar_income)

```

```{r}
# Boxplot of Physical Activity by General Health

## Filtering rows needed for the plot

data_for_plot2 <- data_all %>%
  filter(
    !is.na(pa_total_short_rec),
    !is.na(health_recode),
     ) %>%
  droplevels()  


boxplot_health <- ggplot(
  data_for_plot2,
  aes(x = health_recode, y = pa_total_short_rec, fill = health_recode)
) +
  geom_boxplot(width = 0.7, outlier.shape = NA, alpha = 0.9) +          
  geom_jitter(aes(color = health_recode),                               
              width = 0.15, height = 0, alpha = 0.35, size = 1.7, show.legend = FALSE) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2.8,       
               fill = "white", color = "black", stroke = 0.3) +
  scale_fill_viridis_d(option = "D", end = 0.9, guide = "none") +        
  scale_color_viridis_d(option = "D", end = 0.9, guide = "none") +
  labs(
    title = "Physical Activity by General Health",
    x = "General Health Categories",
    y = "Physical Activity",
    caption = "Boxes = IQR with median; diamond = mean; points = individuals"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey30"),
    axis.text.x   = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    plot.caption  = element_text(color = "grey35", size = 9)
  )

plot(boxplot_health)

```





```{r}

# Heatmap of Mean Physical Activity by Income and Gender

##  Filtering rows needed for the plot
data_for_plot <- data_all %>%
  filter(
    !is.na(pa_total_short_rec),
    !is.na(income_recode),
    !is.na(gender_recode)
  ) %>%
  droplevels()  

## summary 
summary <- data_for_plot %>%
  group_by(income_recode, gender_recode) %>%
  summarise(
    n_total   = n(),
    mean_pa   = mean(pa_total_short_rec, na.rm = TRUE),
    median_pa = median(pa_total_short_rec, na.rm = TRUE),
    sd_pa     = sd(pa_total_short_rec, na.rm = TRUE),
    iqr_pa    = IQR(pa_total_short_rec, na.rm = TRUE),
    .groups   = "drop"
  )


p_heat <- ggplot(summary,
                 aes(x = income_recode, y = gender_recode, fill = mean_pa)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_pa, 1)), size = 3.5) +
  scale_fill_gradientn(
    colours = c("#2166ac", "#f7f7f7", "#b2182b"),
    values  = scales::rescale(c(min(summary$mean_pa, na.rm = TRUE),
                                median(summary$mean_pa, na.rm = TRUE),
                                max(summary$mean_pa, na.rm = TRUE))),
    name = "Mean value"
  ) +
  labs(
    title = "Mean Physical Activity Distribution by Income and Gender",
    x = "Income Categories",
    y = "Gender Categories",
    caption = "Mean Values; rows with missing values removed."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

p_heat

```

